NeuronUnit: Validation Tests for Neurophysiology Models

### 2013-11-04 NeurounitFrontiers/Paper.tex
Added more references.

Test-driven methodologies have started to see success in neuroscience as well. 
Modeling competitions in neuroscience, for example, are typically organized around a collection of simple validation criteria, implemented as executable tests. 
These competitions continue to drive important advances and improve scientists' understanding of the relative merits of different models. 
For example, the quantitative single neuron modeling competition (QSNMC) \citep{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes; 
the ``Hopfield'' challenge \citep{hopfield_what_2000} tested techniques for generating neuronal network form given its function; 
the Neural Prediction Challenge sought the best stimulus reconstructions, given neuronal activity (\url{http://neuralprediction.berkeley.edu}); the Diadem challenge is advancing the art of neurite reconstruction (\url{http://www.diademchallenge.org}); and 
examples from other subfields of biology abound (\url{http://www.the-dream-project.org}).
% \begin{figure}%{r}{0.6\textwidth}
% %\vspace{-38px}
% \includegraphics[scale=0.75]{sciunit_overview.pdf}
% \caption{Proposal overview. 
% \small{i) The \textit{SciUnit} framework can generate discipline-specific libraries for tests generation and model interfacing.  
% \textit{NeuronUnit} is proposed and described here.  Experimental data guides test development; model generation informs and is informed by these \textit{SciUnit} libraries.  
% Each \text{SciUnit} test suite repository on GitHub is indexed by the \textit{SciDash} web application.  
% Test suites for OSB and QSNMC (described in the text) are being built. 
% \textit{SciDash} automatically runs the test collections in each suite and publicly displays the results.}}
% %\vspace{-10px}
% \label{fig:sciunit_overview}
% \end{figure}
% \leavevmode
% \vspace{-0px}
\begin{figure}
\vspace{-38px}
\centering
\includegraphics[scale=0.6]{diagram1.pdf}
\vspace{-55px}
\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the Open Source Brain (OSB) project (\url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (\url{http://neuroelectro.org}).  
Records of test results for various model/test combinations are accessible via SciDash, which indexes GitHub repositories of these records, models, and tests so they can be searched and filtered by the community.}  
\label{fig:sciunit_overview}
\end{figure}
\leavevmode
\vspace{-0px}

Each of these examples has leveraged \emph{ad hoc} infrastructure to support test generation. 
While the specific criteria used to evaluate models varies widely between disciplines in neuroscience, the underlying test-driven methodology has many common features that could be implemented once. 
Recognizing this, we developed a discipline-agnostic framework for developing scientific validation test suites called \textit{SciUnit} (\url{http://www.sciunit.org}). 
Here we describe \textit{NeuronUnit\-}, which builds upon \textit{SciUnit}, allowing neuroscientists to build \textit{SciUnit} tests that validate neurophysiology models against electrophysiological data.  We provide a concrete example pipeline, showing how models described using \text{NeuroML} and provided freely by the Open Source Brain project (\url{http://www.opensourcebrain.org}) can be test in fully automated fashion using published, curated data available through the \textit{NeuroElectro Project} (\url{http://neuroelectro.org}), leveraging facilities from the \textit{NeuroTools} library to extract relevant features of model output.  This is summarized in Figure \ref{fig:sciunit_overview}, which shows the relationships between the layers described here.
% In addition to this open testing toolkit we also provide a library of tests for models of spiking single neurons.

@@

\subsection{\textit{NeuronUnit}: A Suite of Tests for Neurophysiology}\label{sec:neuronunit}
By using machine-readable models from Open Source Brain (OSB), and machine-readable data from resources like The NeuroElectro Project (\url{http://www.neuroelectro.org}), we have produced a suite of data-driven validation tests in \textit{NeuronUnit}, as well as tools for capability implementation, and model interfaces for a representative set of canonical neurophysiology models, all publicly available.

@@

NeuroConstruct (\citep{gleeson_neuroconstruct:_2007}, \url{http://www.neuroconstruct.org}) is a simulation manager that takes NeuroML models and hands off simulation to supported simulators. 
\textit{NeuronUnit} offers a \verbx{sciunit.Model} subclass called \verbx{NeuroConstructModel}, instantiated with the path to a NeuroML model.  
Because NeuroML can describe such a wide range of models, \verbx{NeuroConstructModel} makes few assumptions about them: that each one is \verbx{TimeIntegrable}, and \verbx{HasMembranePotential}.  
It is subclassed to test \textit{specific} NeuroML models (Fig. \ref{fig:ca1_model}).

The Open Source Brain project (OSB, \url{http://www.opensourcebrain.org}) curates many models described in NeuroML. 
OSB-curated projects are converted from their native format into NeuroML, and run on major neural simulators (\url{http://www.neuron.yale.edu/neuron}, \url{http://genesis-sim.org}, \url{http://www.nest-initiative.org}, \url{http://moose.ncbs.res.in}). 
Concordance between model output (beginning with the NeuroML description) and reference output (from native simulator source files) is reported for each model. 
Thus, OSB is an excellent source of models that, in addition to being open source, are sufficiently described to enable validation. 
The hippocampal CA1 pyramidal cell is commonly modeled, and we implement one such model hosted on OSB (\url{http://opensourcebrain.org/projects/ca1pyramidalcell}) by simply declaring a \verbx{CA1PyramidalCellModel} class, inheriting from \verbx{NeuroConstructModel}.  
This basic implementation simply ``wraps'' the components of the existing model, with simulator interaction taken care of by \verbx{NeuroConstructModel} methods; 
thus, only the code shown in Fig. \ref{fig:ca1_model} is required. 
All OSB models, and indeed any NeuroML model, can be tested similarly.

@@

\subsection{Participation from Modeling Communities}
Modelers may not want to expose model capabilities, a requirement of test-taking.  
We anticipate four solutions: \textbf{First}, interfacing a model to \textit{SciUnit} requires only implementing selected model capabilities. 
Often this means identifying native model procedures that satisfy a capability, and wrapping their functionality. 
This can require as little as one line of code. 
Importantly, the modeler is not required to expose or rewrite any model flow control. 
\textbf{Second}, we support multiple environments automatically by using NeuroML \citep{gleeson_neuroml:_2010}, and other simulator-independent model descriptions are possible for other domains. 
Automated generation of NeuroML from native model source code is in development (Gleeson, personal communication); for the popular NEURON simulator (\url{http://www.neuron.yale.edu/neuron}), this functionality is already mature and in use. 
This minimizes modeler effort for a large and growing number of models. 
\textbf{Third}, modelers have an incentive to demonstrate publicly their models' validity. 
Participation in public modeling competitions demonstrates this incentive. 
\textbf{Fourth}, modelers have an incentive to use \textit{SciUnit} during development (see TDD, above) to ensure that ongoing development preserves correspondence between model and data. 
A popular test suite can represent a ``gold standard'' by which progress during development is judged.

@@

\subsection{Diversity of Levels and Kinds of Models and Data}
The diversity of topics in biology is vast. 
\textbf{First}, we address this by providing an interface allowing modelers to express specific capabilities.  
This capability set determines the range of eligible tests.  Scale hierarchies are embedded in capability inheritance.  =
For example, \verbx{HasActionPotentials} inherits from \verbx{HasNeurons}, and \verbx{HodgkinHuxley} inherits from \verbx{VoltageGated}. 
Thus, incompatibility of a test-requiring-action-potentials for a model-lacking-neurons is known without explicit tagging. 
\textbf{Second}, NeuroML naturally addresses diversity of scales because it is organized hierarchically, in ``levels.''  
Models can be sub- or supersets of other models; similarly for SBML (\citep{hucka_systems_2003}, \url{http://sbml.org}) a general systems biology markup language. 
\textbf{Third}, cross-level testing can use ``Representional Similarity Analysis'' (RSA) \citep{kriegeskorte_representational_2008}, requiring only that a model respond to defined inputs (e.g. stimuli). 
A ``similarity matrix'' for input responses defines a unique model signature, and can serve as intermediate test output. 
Goodness-of-fit between similarity matrices for model and experiment determines test scores; 
these matrices are independent of model scale because their size depends only on test inputs, not system detail.

\subsection{Arbitrary Scoring Criteria for Tests}
A test first assesses goodness-of-fit, and applies a normalization (e.g. pass/fail, 0.0-1.0) to generate a score. 
Arbitrary choices at both stages may benefit some models over others.  \textbf{First}, however, rank-ordering is constant across many goodness-of-fit metrics, meaning that choice of metric will rarely cause an otherwise passing model to fail and vice versa. 
For example, given a data mean and variance, ordering model output by Z-score or p-value will yield the same relative ranking of models. 
Indeed, rank ordering of models may prove more valuable than test scores themselves. 
\textbf{Second}, suite repositories are open (e.g. Fig. \ref{fig:scidash_repo}), so tests can be cloned and new statistical choices implemented. 
Statistics as applied to neuroscience have been notoriously ``fast and loose''; 
identification and correction of flawed methodology is becoming increasingly common (\citep{button_power_2013,kriegeskorte_circular_2009,galbraith_study_2010}, \url{http://prefrontal.org/files/posters/Bennett-Salmon-2009.jpg}), and is accelerated by an open framework. 
The validation criteria is made explicit in the specification of a test, so modelers need not guess which criteria are being used to validate or invalidate their model. 
Validation criteria are subject to debate (indeed, the QSNMC criteria changed between 2007 and 2008 due to such debates), and neuroscientists who wish to promote different criteria need only derive alternative tests. 
A community consensus will slowly emerge in the form a commonly-accepted test suite (see Sec. \ref{sec:scidash}). 
This community can judge which test version is most appropriate, i.e. what a model \textit{should} do -- this process documented via standard moderation techiniques used on GitHub -- and the \textit{SciUnit} framework determines whether the model \textit{does} it.

@@

\subsection{Expansion Into Other Areas of Biology}
After covering neurophysiology, we would like \textit{SciUnit} to be applied across neuroscience and in other biological sciences. 
The framework is discipline-agnostic, so community participation and model description are the only obstacles. 
Community participation begins with enumerating the capabilities relevant to a sub-discipline, and then writing tests. 
Model description can expand within NeuroML (which already covers multiple levels within neuroscience) and tools for capability implementation can incorporate libraries for neuroimaging (NiBabel, \url{http://nipy.org/nibabel}), neuroanatomy (NeuroHDF, \url{http://neurohdf.readthedocs.org}) and other sub-disciplines. 
SBML (\citep{hucka_systems_2003}, \url{http://sbml.org} will enable expansion beyond neuroscience, facilitated by parallel efforts among NeuroML developers to interface with it (Crook, unpublished). 
One intriguing possiblity is applying \textit{SciUnit} to the OpenWorm project (\url{http://www.openworm.org}), which through open, collaborative development seeks to model the entire organism C. elegans.

>>>>

Test-driven methodologies have started to see success in neuroscience as well. 
Modeling competitions in neuroscience, for example, are typically organized around a collection of simple validation criteria, implemented as executable tests. 
These competitions continue to drive important advances and improve scientists' understanding of the relative merits of different models. 
For example, the quantitative single neuron modeling competition (QSNMC) \citep{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes; 
the ``Hopfield'' challenge \citep{hopfield_what_2000} tested techniques for generating neuronal network form given its function; 
the Neural Prediction Challenge sought the best stimulus reconstructions, given neuronal activity (\url{http://neuralprediction.berkeley.edu}); the Diadem challenge is advancing the art of neurite reconstruction (\url{http://www.diademchallenge.org}); and 
examples from other subfields of biology abound (\url{http://www.the-dream-project.org}).
% \begin{figure}%{r}{0.6\textwidth}
% %\vspace{-38px}
% \includegraphics[scale=0.75]{sciunit_overview.pdf}
% \caption{Proposal overview. 
% \small{i) The \textit{SciUnit} framework can generate discipline-specific libraries for tests generation and model interfacing.  
% \textit{NeuronUnit} is proposed and described here.  Experimental data guides test development; model generation informs and is informed by these \textit{SciUnit} libraries.  
% Each \text{SciUnit} test suite repository on GitHub is indexed by the \textit{SciDash} web application.  
% Test suites for OSB and QSNMC (described in the text) are being built. 
% \textit{SciDash} automatically runs the test collections in each suite and publicly displays the results.}}
% %\vspace{-10px}
% \label{fig:sciunit_overview}
% \end{figure}
% \leavevmode
% \vspace{-0px}
\begin{figure}
\vspace{-38px}
\centering
\includegraphics[scale=0.6]{diagram1.pdf}
\vspace{-55px}
\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the \textit{Open Source Brain Project} (OSB, \citep{gleeson_open_2012}, \url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (Neuroelectro, \citep{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}).  
Records of test results for various model/test combinations are accessible via SciDash, which indexes GitHub repositories of these records, models, and tests so they can be searched and filtered by the community.}  
\label{fig:sciunit_overview}
\end{figure}
\leavevmode
\vspace{-0px}

Each of these examples has leveraged \emph{ad hoc} infrastructure to support test generation. 
While the specific criteria used to evaluate models varies widely between disciplines in neuroscience, the underlying test-driven methodology has many common features that could be implemented once. 
Recognizing this, we developed a discipline-agnostic framework for developing scientific validation test suites called \textit{SciUnit} (\url{http://www.sciunit.org}). 
Here we describe \textit{NeuronUnit\-}, which builds upon \textit{SciUnit}, allowing neuroscientists to build \textit{SciUnit} tests that validate neurophysiology models against electrophysiological data.  We provide a concrete example pipeline, showing how models described using \text{NeuroML} and provided freely by the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}) can be tested in fully automated fashion using published, curated data available through the \textit{NeuroElectro Project} (Neuroelectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}), leveraging facilities from the \textit{NeuroTools} library (\url{http://neuralensemble.org/NeuroTools}) to extract relevant features of model output.  This is summarized in Figure \ref{fig:sciunit_overview}, which shows the relationships between the layers described here.
% In addition to this open testing toolkit we also provide a library of tests for models of spiking single neurons.

@@

\subsection{\textit{NeuronUnit}: A Suite of Tests for Neurophysiology}\label{sec:neuronunit}
By using machine-readable models from \textit{OSB}, and machine-readable data from resources like \textit{NeuroElectro}, we have produced a suite of data-driven validation tests in \textit{NeuronUnit}, as well as tools for capability implementation, and model interfaces for a representative set of canonical neurophysiology models, all publicly available.

@@

NeuroConstruct (\cite{gleeson_neuroconstruct:_2007}, \url{http://www.neuroconstruct.org}) is a simulation manager that takes NeuroML models and hands off simulation to supported simulators. 
\textit{NeuronUnit} offers a \verbx{sciunit.Model} subclass called \verbx{NeuroConstructModel}, instantiated with the path to a NeuroML model.  
Because NeuroML can describe such a wide range of models, \verbx{NeuroConstructModel} makes few assumptions about them: that each one is \verbx{TimeIntegrable}, and \verbx{HasMembranePotential}.  
It is subclassed to test \textit{specific} NeuroML models (Fig. \ref{fig:ca1_model}).

\textit{OSB} curates many models described in NeuroML. 
OSB-curated projects are converted from their native format into NeuroML, and run on major neural simulators such as Neuron (\cite{carnevale_neuron_2006}, \url{http://www.neuron.yale.edu/neuron}), GENESIS (\cite{bower_genesis_2007}, \url{http://genesis-sim.org}), NEST (\cite{gewaltig_nest_2007}, \url{http://www.nest-initiative.org}), and MOOSE (\cite{ray_moose_2008}, \url{http://moose.ncbs.res.in}). 
Concordance between model output (beginning with the NeuroML description) and reference output (from native simulator source files) is reported for each model. 
Thus, OSB is an excellent source of models that, in addition to being open source, are sufficiently described to enable validation. 
The hippocampal CA1 pyramidal cell is commonly modeled, and we implement one such model hosted on OSB (\url{http://opensourcebrain.org/projects/ca1pyramidalcell}) by simply declaring a \verbx{CA1PyramidalCellModel} class, inheriting from \verbx{NeuroConstructModel}.  
This basic implementation simply ``wraps'' the components of the existing model, with simulator interaction taken care of by \verbx{NeuroConstructModel} methods; 
thus, only the code shown in Fig. \ref{fig:ca1_model} is required. 
All OSB models, and indeed any NeuroML model, can be tested similarly.

@@

\subsection{Participation from Modeling Communities}
Modelers may not want to expose model capabilities, a requirement of test-taking.  
We anticipate four solutions: \textbf{First}, interfacing a model to \textit{SciUnit} requires only implementing selected model capabilities. 
Often this means identifying native model procedures that satisfy a capability, and wrapping their functionality. 
This can require as little as one line of code. 
Importantly, the modeler is not required to expose or rewrite any model flow control. 
\textbf{Second}, we support multiple environments automatically by using NeuroML \citep{gleeson_neuroml:_2010}, and other simulator-independent model descriptions are possible for other domains. 
Automated generation of NeuroML from native model source code is in development (Gleeson, personal communication); for the popular NEURON simulator (\cite{carnevale_neuron_2006}), this functionality is already mature and in use. 
This minimizes modeler effort for a large (at least 1000, \url{http://www.neuron.yale.edu/neuron/node/69}) and growing number of neuronal models. 
\textbf{Third}, modelers have an incentive to demonstrate publicly their models' validity. 
Participation in public modeling competitions demonstrates this incentive. 
\textbf{Fourth}, modelers have an incentive to use \textit{SciUnit} during development (see TDD, above) to ensure that ongoing development preserves correspondence between model and data. 
A popular test suite can represent a ``gold standard'' by which progress during development is judged.

@@

\subsection{Diversity of Levels and Kinds of Models and Data}
The diversity of topics in biology is vast. 
\textbf{First}, we address this by providing an interface allowing modelers to express specific capabilities.  
This capability set determines the range of eligible tests.  Scale hierarchies are embedded in capability inheritance.  =
For example, \verbx{HasActionPotentials} inherits from \verbx{HasNeurons}, and \verbx{HodgkinHuxley} inherits from \verbx{VoltageGated}. 
Thus, incompatibility of a test-requiring-action-potentials for a model-lacking-neurons is known without explicit tagging. 
\textbf{Second}, NeuroML naturally addresses diversity of scales because it is organized hierarchically, in ``levels.''  
Models can be sub- or supersets of other models; similarly for SBML (\cite{hucka_systems_2003}, \url{http://sbml.org}) a general systems biology markup language. 
\textbf{Third}, cross-level testing can use ``Representional Similarity Analysis'' (RSA) \citep{kriegeskorte_representational_2008}, requiring only that a model respond to defined inputs (e.g. stimuli). 
A ``similarity matrix'' for input responses defines a unique model signature, and can serve as intermediate test output. 
Goodness-of-fit between similarity matrices for model and experiment determines test scores; 
these matrices are independent of model scale because their size depends only on test inputs, not system detail.

\subsection{Arbitrary Scoring Criteria for Tests}
A test first assesses goodness-of-fit, and applies a normalization (e.g. pass/fail, 0.0-1.0) to generate a score. 
Arbitrary choices at both stages may benefit some models over others.  \textbf{First}, however, rank-ordering is constant across many goodness-of-fit metrics, meaning that choice of metric will rarely cause an otherwise passing model to fail and vice versa. 
For example, given a data mean and variance, ordering model output by Z-score or p-value will yield the same relative ranking of models. 
Indeed, rank ordering of models may prove more valuable than test scores themselves. 
\textbf{Second}, suite repositories are open (e.g. Fig. \ref{fig:scidash_repo}), so tests can be cloned and new statistical choices implemented. 
Statistics as applied to neuroscience have been notoriously ``fast and loose''; 
identification and correction of flawed methodology is becoming increasingly common (\cite{button_power_2013,kriegeskorte_circular_2009,galbraith_study_2010}, \url{http://prefrontal.org/files/posters/Bennett-Salmon-2009.jpg}), and is accelerated by an open framework. 
The validation criteria is made explicit in the specification of a test, so modelers need not guess which criteria are being used to validate or invalidate their model. 
Validation criteria are subject to debate (indeed, the QSNMC criteria changed between 2007 and 2008 due to such debates), and neuroscientists who wish to promote different criteria need only derive alternative tests. 
A community consensus will slowly emerge in the form a commonly-accepted test suite (see Sec. \ref{sec:scidash}). 
This community can judge which test version is most appropriate, i.e. what a model \textit{should} do -- this process documented via standard moderation techiniques used on GitHub -- and the \textit{SciUnit} framework determines whether the model \textit{does} it.  This process can be made even more visible by storing the output of test runs using Sage or IPython worksheets, or through optional database backends such as in the Sumatra lab notebook for simulation and analysis (\cite{sumatra_davison_2012}, \url{http://neuralensemble.org/sumatra}).

@@

\subsection{Expansion Into Other Areas of Biology}
After covering neurophysiology, we would like \textit{SciUnit} to be applied across neuroscience and in other biological sciences. 
The framework is discipline-agnostic, so community participation and model description are the only obstacles. 
Community participation begins with enumerating the capabilities relevant to a sub-discipline, and then writing tests. 
Model description can expand within NeuroML (which already covers multiple levels within neuroscience) and tools for capability implementation can incorporate libraries for neuroimaging (NiBabel, \url{http://nipy.org/nibabel}), neuroanatomy (NeuroHDF, \url{http://neurohdf.readthedocs.org}) and other sub-disciplines. 
SBML will enable expansion beyond neuroscience, facilitated by parallel efforts among NeuroML developers to interface with it (Crook, unpublished). 
One intriguing possiblity is applying \textit{SciUnit} to the OpenWorm project (\url{http://www.openworm.org}), which through open, collaborative development seeks to model the entire organism C. elegans.
